---
title: "Wine Analysis"
author: "Jack Leary"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
# knitr::opts_knit$set(root.dir = "Users/Jack/Documents/My-Projects")
```

```{r, message=FALSE, warning = FALSE}
# libraries
library(tidyverse)
library(plotly)
library(e1071)
library(randomForest)
library(caret)
library(keras)
library(xgboost)
```

```{r, message=FALSE}
# load data
wineFull <- read_csv("/Users/Jack/Desktop/Datasets/winequalityN.csv")
```

After doing some real basic data cleaning, I centered and scaled each numerical feature so that I could overlay a $N(0, 1)$ density line over each plot to check for normality, skew, and kurtosis within each feature.
```{r}
# cleaning
wineFull$type <- as.factor(wineFull$type)
wineFull$quality <- as.factor(wineFull$quality)
colnames(wineFull) <- c("type", "fixedAcidity", "volatileAcidity", "citricAcid", 
                        "residualSugar", "chlorides", "freeSulfurDioxide", 
                        "totalSulfurDioxide", "density", "pH", "sulphates", 
                        "percentAlcohol", "quality")
wineFullCopy <- wineFull
wineFullCopy[, 2:12] <- scale(wineFullCopy[, 2:12], center = TRUE, scale = TRUE)
wineFull <- na.omit(wineFull)
wineFullCopy <- na.omit(wineFullCopy)
```

From the histograms, it appears that volatile acidity and residual sugar are both skewed right, while total sulfur dioxide looks bimodal. The other variables seem approximately normally distributed, and I don't believe there's sufficient reason to transform any of the features further.
```{r, echo=FALSE, warning=FALSE}
# setting up histograms to check for normality/kurtosis
set.seed(629)
x <- rnorm(1000000)
fit <- density(x)
x2 <- rnorm(1000000, mean = 6, sd = 1)
fit2 <- density(x2)
p1 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~fixedAcidity, name = "Histogram") %>% 
  layout(title = "Fixed Acidity Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p2 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~volatileAcidity) %>% 
  layout(title = "Volatile Acidity Histogram", name = "Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p3 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~citricAcid, name = "Histogram") %>% 
  layout(title = "Citric Acid Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p4 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~residualSugar, name = "Histogram") %>% 
  layout(title = "Residual Sugar Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p5 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~chlorides, name = "Histogram") %>% 
  layout(title = "Chlorides Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p6 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~freeSulfurDioxide, name = "Histogram") %>% 
  layout(title = "Free Sulfur Dioxide Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p7 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~totalSulfurDioxide, name = "Histogram") %>% 
  layout(title = "Total Sulfur Dioxide Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p8 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~density, name = "Histogram") %>% 
  layout(title = "Density Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p9 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~pH, name = "Histogram") %>% 
  layout(title = "pH Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p10 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~sulphates, name = "Histogram") %>% 
  layout(title = "Sulphates Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p11 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~percentAlcohol, name = "Histogram") %>% 
  layout(title = "Alcohol by Volume Histogram") %>% 
  add_trace(x = fit$x, y = fit$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(0, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))
p12 <- plot_ly(wineFullCopy) %>% 
  add_histogram(x = ~quality, name = "Histogram") %>% 
  layout(title = "Quality Histogram") %>% 
  add_trace(x = fit2$x, y = fit2$y, type = "scatter", mode = "lines", fill = NULL, 
            yaxis = "y2", name = "N(6, 1) Density") %>% 
  layout(yaxis2 = list(overlaying = "y", side = "right"))

p1
p2
p3
p4
p5
p6
p7
p8
p9
p10
p11
p12
```

I originally thought that I wouldn't be able to include alohol percentage as a feature. I believed that alcohol percentage didn't differ in between red and white wines, thus including it would be statistical malpractice. However, after running a two-sided t-test, it appears that there is a small but highly significant difference between the two. 
```{r}
redAlc <- wineFull[wineFull$type == "red", ]$percentAlcohol
whiteAlc <- wineFull[wineFull$type == "white", ]$percentAlcohol
t.test(redAlc, whiteAlc, alternative = "two.sided", paired = FALSE)
```

```{r}
# train/test set creation
set.seed(629)
trainIdx <- sample(seq_len(nrow(wineFull)), size = nrow(wineFull) * .75)
wineTrain <- wineFull[trainIdx, ]
wineTest <- wineFull[-trainIdx, ]
```

I decided to use a random forest to select which features to include in modeling later. From the variable importance plot, it looks like totalSulfurDioxide, chlorides, volatileAcidity, sulphates, and density are the 5 most important variables for reducing error, so I'll start with those and adjust as needed.
```{r randomForest}
# random forest
rf1 <- randomForest(x = wineTrain[, 2:12], y = wineTrain$type, 
                    ntree = 500, importance = TRUE)
varImpPlot(rf1)
```

```{r binomialGLM}
# binomial glm
glm1 <- glm(type ~ ., data = wineTrain, family = "binomial")
summary(glm1) # resid deviance = 307.43, AIC = 343.43

glm2 <- glm(type ~ totalSulfurDioxide + chlorides + volatileAcidity + sulphates + density, 
            data = wineTrain, family = "binomial")
summary(glm2) # resid deviance = 558.35, AIC = 570.35

glm3 <- glm(type ~ volatileAcidity + residualSugar + chlorides + freeSulfurDioxide + 
              totalSulfurDioxide + density + sulphates + percentAlcohol, data = wineTrain, 
            family = "binomial")
summary(glm3) # resid deviance = 331.18, AIC = 349.18
predictGLM <- vector()
probsGLM <- predict(glm3, wineTest, type = "response")
for(i in 1:length(probsGLM)) {
  if(probsGLM[i] > .5) {
    predictGLM[i] <- "white"
  } else {
    predictGLM[i] <- "red"
  }
}
predictGLM <- as.factor(predictGLM)
confusionMatrix(wineTest$type, predictGLM)
```

```{r svm}
set.seed(629)
tune1 <- tune(svm, type ~ volatileAcidity + residualSugar + chlorides + freeSulfurDioxide + 
              totalSulfurDioxide + density + sulphates + percentAlcohol, data = wineTrain, 
              ranges = list(gamma = 10^(-5:-1), cost = 10^(-3:1)))
tune1$best.parameters
tune1$best.model

svm1 <- svm(type ~ volatileAcidity + residualSugar + chlorides + freeSulfurDioxide + 
              totalSulfurDioxide + density + sulphates + percentAlcohol, data = wineTrain, 
              kernel = "radial", gamma = 0.1, cost = 10)
predictSVM <- predict(svm1, wineTest)
confusionMatrix(wineTest$type, predictSVM)
```

```{r XGBoost}
set.seed(629)
xgb1 <- train(type ~ volatileAcidity + residualSugar + chlorides + freeSulfurDioxide + 
              totalSulfurDioxide + density + sulphates + percentAlcohol, data = wineTrain, 
              method = "xgbTree", trControl = trainControl("cv", number = 10))
xgb1$bestTune
predictXGB <- xgb1 %>% predict(wineTest)
confusionMatrix(wineTest$type, predictXGB)
```







