---
title: "Downstream Analysis of Human Single Cell RNAseq Data Using Seurat"
author: "Jack Leary"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: yeti
    highlight: tango
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Introduction
This a a downstream analysis pipeline based on the `Seurat` package developed and maintained by the Satija Lab at the New York Genome Center. More information on the Satija Lab, their research, and the `Seurat` package can be found [here](https://satijalab.org). 

```{r libraries}
library(dplyr)
library(Seurat)
library(DESeq2)
library(SingleR)
library(scRNAseq)
library(SingleCellExperiment)
library(scater)
library(decoderr)
library(uwot)
library(phateR)
library(reticulate)
library(jackknife)
library(DT)
library(sctransform)
```

Here we import the results of the `cellranger` analysis pipeline and save them in a Seurat object. Upon creation of the Seurat object, the sparse counts are subsetted to only include those genes that are found in at least three cells, and only those cells that include information on at least two hundred genes.
```{r import}
path1 <- "/Volumes/Home/Jen Jen Yeh Lab/YeSu/Gen_RData/Single-cell/Moncada_GSE111672/"
samplename <- "PDAC_B"
path2 <- paste0(path1, samplename, "_outs")
sparse_counts <- Read10X(data.dir = path2)
sc <- CreateSeuratObject(counts = sparse_counts, 
                         project = samplename, 
                         min.cells = 3, 
                         min.features = 200)
```

A peek at the imported counts shows us that they are saved in a sparse matrix, a specialized file format that saves space by representing zero values as "$.$", which is necessary because single cell RNAseq counts are composed mostly of zero values. This is a unique problem in the data analysis of scRNAseq data that we will address later.
```{r peek}
sparse_counts[c("AURKA", "TP53", "BRCA2"), 1:4]
```

# Quality Control
Some quality control metrics are computed automatically when the Seurat object is created. Let's take a look. 
```{r init_qc}
datatable(sc@meta.data, filter = "top", options = list(pageLength = 5, scrollX = TRUE))
```

## Mitochondrial QC
Low quality cells often map a large percentage of their reads to the mitochondrial genome. Here we'll calculate that percentage for each cell.
```{r mito}
sc[["percent_MT"]] <- PercentageFeatureSet(sc, pattern = "^MT-")
datatable(sc@meta.data, filter = "top", options = list(pageLength = 5, scrollX = TRUE))
```

Here we visually compare different QC metrics, including the percentage of mitochondrial DNA that we just added.
```{r mito_viz}
VlnPlot(sc, features = c("nFeature_RNA", "nCount_RNA", "percent_MT"), ncol = 3, 
        pt.size = .1)
```

## Normalization and Scaling
In this step, we use the `SCTransform` wrapper provided in `Seurat` to regress out unwanted sources of variation in our data (such as the percentage of mitochondrial DNA), as well as normalize and scale it. The function has two goals: to remove correlations of gene expression with sequencing depth, and to preserve biological heterogeneity while removing technical noise. This is accomplished by using a regularized negative binomial GLM to model the variance of each gene, and then normalizing its expression using the Pearson residuals of the model. `SCTransform()` also automatically selects the 3,000 most variable genes, based on the GLM, upon which we will perform further downstream analysis.
```{r sctransform}
sc <- SCTransform(sc, vars.to.regress = "percent_MT", seed.use = 629, verbose = FALSE)
```

# Dimension Reduction
## Scaling
Scaling and centering the data is a necessary part of principal component analysis (PCA), which will be the building block for the rest of our dimension reduction and cluster analysis. If this transformation prove ineffective, we can make use of the `sctransform` workflow provided by the Satija Lab to regress out unwanted sources of variation. 
```{r scale}
sc <- ScaleData(sc)
```

## PCA
When running PCA, we'll use only the 2000 highly variable genes we identified earlier. This speeds up the process, and ensures that we're capturing sources of variation that are potentially biologically interesting. The visualization techniques below help us to determine sources of heterogeneity in our data.
```{r pca, results='hide'}
sc <- RunPCA(sc, features = VariableFeatures(sc), npcs = 50)
VizDimLoadings(sc, dims = 1:2, reduction = "pca")
DimPlot(sc, reduction = "pca")
DimHeatmap(sc, dims = 1:6, cells = 1500, balanced = TRUE, reduction = "pca")
```

## Assigning Cell Cycle Scores
Here we assign a stage of the cell cycle to each of our cells, based on a reference set of periodically expressed genes taken from Tirosh *et al* (2015). This isn't a dimensionality reduction step *per se*, but it's necessary to run the cell scoring algorithm in PCA space. We can later use the results to determine whether or not our clusters are determined by cell cycle stage rather than cell identity, and if so we can regress out the cell cycle effect using a generalized linear model.
```{r cycle_scores}
s_genes <- cc.genes.updated.2019$s.genes
g2m_genes <- cc.genes.updated.2019$g2m.genes
sc <- CellCycleScoring(sc, s.features = s_genes, g2m.features = g2m_genes, set.ident = FALSE)
datatable(sc@meta.data, filter = "top", options = list(pageLength = 5, scrollX = TRUE))
```

### Choosing principal components
PCA is an unsupervised technique; this implies that just because it identifies a source of variation, that variation might just be noise. Fortunately, several techniques exist to aid us in determining which PCs capture statistically significant sources of heterogeneity. The `JackStraw` function uses bootstrap sampling and PCA to produce p-values for each principal component. We can see via the plot that the significance of the PCs stays strong through PC 50, so we'll use all 50 PCs going forward.
```{r pca_check, results='hide'}
sc <- JackStraw(sc, reduction = "pca", num.replicate = 100, dims = 50)
sc <- ScoreJackStraw(sc, dims = 1:50)
JackStrawPlot(sc, dims = 1:50)
```

## Assigning Cell Cycle Scores
Here we assign a stage of the cell cycle to each of our cells, based on a reference set of periodically expressed genes taken from Tirosh *et al* (2015). This isn't a dimensionality reduction step per se, but it's necessary to run the cell scoring algorithm in PCA space. We can later use the results to determine whether or not our clusters are determined by cell cycle stage rather than cell identity, and if so we can regress out the cell cycle effect using a generalized linear model.
```{r cycle_scores}
s_genes <- cc.genes.updated.2019$s.genes
g2m_genes <- cc.genes.updated.2019$g2m.genes
sc <- CellCycleScoring(sc, s.features = s_genes, g2m.features = g2m_genes, set.ident = FALSE)
datatable(sc@meta.data, filter = "top", options = list(pageLength = 5, scrollX = TRUE))
```

## t-SNE
t-Distributed Stochastic Neighbors Embedding is a non-linear dimension reduction technique that attempts to preserve both the global and local structure of the input data. It uses gradient descent to recreate the high-dimensional probability distribution of the points in a low-dimensional space. The perplexity parameter of t-SNE determines how much you'd like to preserve local structure over global structure, but it's default is set to 30. It's a good idea to provide a seed to the t-SNE algorithm since, due to the nature of the non-convex optimization problem it attempts to solve, you can receive different results from different runs. Setting a seed ensures that the algorithm always starts at the same point.
```{r tsne}
sc <- RunTSNE(sc, reduction = "pca", seed.use = 629, dims = 1:45, dim.embed = 2, check_duplicates = FALSE)
```

## UMAP
Uniform Manifold Approximation and Projection is a fairly recent development in the area of dimension reduction that was proposed by McInnes *et al.* in 2018. It is very similar to t-SNE, although it's authors claim that it preserves both the global and local structure of your data. This means that the distances between the clusters in the low-dimensional space actually represent the distances between the clusters in the high-dimensional space, which they do not in t-SNE. Like t-SNE, it is a non-deterministic method, and it is best practice to provide a seed. 
```{r umap, results='hide'}
sc <- RunUMAP(sc, reduction = "pca", dims = 1:45, umap.method = "uwot", 
              n.components = 2, seed.use = 629, n.epochs = 700)
```

## PHATE
Potential of Heat-diffusion for Affinity-based Trajectory Embedding is a very new method published by Moon *et al* at the end of 2019. Like UMAP, the authors claim that the algorithm "captures both local and global nonlinear structure." The algorithm was produced specifically with the analysis of high throughput sequencing data, such as single cell RNAseq, in mind.We'll need to use Python to implement the PHATE algorithm.
```{r}
pca_df <- data.frame(sc@reductions$pca@cell.embeddings)
```

We import the Python libraries we'll need to run PHATE.
```{python}
import pandas as pd
import phate
```

After importing the necessary libraries, we initialize an instance of the PHATE function and fit the algorithm to the data. 
```{python}
pca_df = r.pca_df
phate_op = phate.PHATE(n_components=2, knn=5, mds_solver='sgd', random_state=629, n_pca=None)
phate_df = phate_op.fit_transform(pca_df)
```

Next we call the results from Python back into R, and add them to our Seurat object.
```{r phate}
phate_df <- as.matrix(py$phate_df)
rownames(phate_df) <- colnames(sc)
phate_obj <- CreateDimReducObject(embeddings = phate_df, 
                                  key = "PHATE_", 
                                  assay = "RNA", 
                                  global = TRUE)
sc@reductions$phate <- phate_obj
```

# Clustering
The main clustering method implemented in `Seurat` is graph-based. This type of clustering makes no assumptions about the true number of clusters, or the shape or density of the data, which is useful because scRNAseq data is so sparse. Cells are embedded onto a k-nearest neighbors graph in PCA space, and then iteratively sorted into clusters.
```{r graph, results='hide'}
sc <- FindNeighbors(sc, reduction = "pca")
sc <- FindClusters(sc, resolution = .5)
```

## Visualization
After performing clustering, we can use the dimension reduction results we produced earlier to visualize our clusters. We can see that PCA actually does an OK job separating the clusters, although non-linear methods will certainly improve the visualization. The second PCA plot shows that cell cycle stage is not a main driver of heterogeneity, at least in PCA space.
```{r pca_plot}
DimPlot(sc, reduction = "pca", dims = c(1, 2))
DimPlot(sc, reduction = "pca", dims = c(1, 2), group.by = "Phase")
```

t-SNE shows decent separation between clusters, and we can see once again that cell cycle stage doesn't appear to be affecting our clusters.
```{r tsne_plot}
DimPlot(sc, reduction = "tsne", dims = c(1, 2))
DimPlot(sc, reduction = "tsne", dims = c(1, 2), group.by = "Phase")
```

We can see here that UMAP does the best job (so far) of clearly separating our clusters.
```{r umap_plot}
DimPlot(sc, reduction = "umap", dims = c(1, 2))
```

PHATE is used to both separate clusters while preserving global and local structure and identify biological processes in the form of trajectories. The PHATE plot shows interesting potential trajectories in clusters 5, 7, and 9.
```{r phate_plot}
DimPlot(sc, reduction = "phate", dims = c(1, 2))
DimPlot(sc, reduction = "phate", dims = c(1, 2), group.by = "Phase")
```

At this point, it's a good idea to save and reload the Seurat object so that we don't need to perform all the previous steps each time we want to do more downstream analysis.
```{r, results='hide'}
saveRDS(sc, file = paste0("/Volumes/Home/Jen Jen Yeh Lab/Jack/scRNAseq/Seurat/RData Files/Moncada_",
                          samplename, "_hg38.Rds"))
sc <- readRDS(paste0("/Volumes/Home/Jen Jen Yeh Lab/Jack/scRNAseq/Seurat/RData Files/Moncada_",
                          samplename, "_hg38.Rds"))
```

# Biomarker Identification
After clustering and visualizing our cells, we'd like to assign cell types to each cluster. This is accomplished by determining which genes are significantly differentially expressed between clusters, and comparing those genes to canonical markers. 
```{r}
sc_markers <- FindAllMarkers(sc, only.pos = TRUE, test.use = "wilcox")
top_markers <- sc_markers %>% group_by(cluster) %>% top_n(n = 1, wt = avg_logFC)
top_markers
```

## Visualization
We visualize these results first in the context of the UMAP dimension reduction plot. 
```{r}
for (i in seq(nrow(top_markers))) {
  print(FeaturePlot(sc, features = top_markers$gene[i], reduction = "umap", dims = c(1, 2)))
}
```

Next, we use a violin plot of gene expression grouped by cluster to show how much each marker is differentially expressed.
```{r}
for (i in seq(nrow(top_markers))) {
  p3 <- VlnPlot(sc, features = top_markers$gene[i])
  print(p3)
}
```

Finally, we plot a heatmap of the top five marker genes per cluster, showing that there are in fact significant differences in gene expression levels between clusters, with the possible exception of cluster 0. 
```{r}
top5 <- sc_markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_logFC)
DoHeatmap(sc, features = top5$gene, angle = 45) + NoLegend()
```

## Assigning Cell Type Identities
Here we use the `SingleR` package developed by Lun *et al* to assign cell identities to the clusters previously generated by `Seurat`. At the moment, differential expression analysis is performed within `SingleR` using a Wilcox test. Once the kinks are worked out concerning `DESeq2` in the `FindAllMarkers()` function, we can provide `SingleR` with our own marker genes. `SingleR` uses reference data to assess similarity between our cell clusters and labeled expression data. The bulk reference data is comprised of microarrays from Mabbott *et al* (2013), whereas the single cell sequencing data comes from Baron *et al* (2017). Previously, the single cell reference was run through the same `sctransform` normalization workflow as our sample. 
```{r SingleR}
norm_counts <- data.frame(sc@assays$RNA@data)
bulk_ref <- HumanPrimaryCellAtlasData()
sc_ref <- readRDS("/Volumes/Home/Jen Jen Yeh Lab/Jack/scRNAseq/Seurat/single_cell_ref_normalized.Rds")
sc_ref <- sc_ref[, !is.na(sc_ref$label)]

bulk_preds <- SingleR(test = norm_counts, ref = bulk_ref, labels = bulk_ref$label.main, 
                      method = "cluster", clusters = sc$seurat_clusters, de.method = "wilcox")
table(bulk_preds$labels)

sc_preds <- SingleR(test = norm_counts, ref = sc_ref, labels = sc_ref$label, 
                 method = "cluster", clusters = sc$seurat_clusters, de.method = "wilcox")
table(sc_preds$labels)

sc[["SingleR.labels.bulk"]] <- bulk_preds$labels[match(sc[[]][["seurat_clusters"]], 
                                                          rownames(bulk_preds))]
sc[["SingleR.labels.sc"]] <- sc_preds$labels[match(sc[[]][["seurat_clusters"]], 
                                                   rownames(sc_preds))]
```

According to the bulk reference, this sample has several epithelial clusters, inidcating that we have several types of epithelial cell. This was to be expected as the results of Moncada *et al* (2020) showed several ductal subclusters. The PHATE plot shows interesting trajectories in the endothelial, T, and neutrophil clusters.
```{r}
DimPlot(sc, reduction = "tsne", label = TRUE, group.by = "SingleR.labels.bulk", repel = TRUE)
DimPlot(sc, reduction = "tsne", label = TRUE, group.by = "SingleR.labels.sc", repel = TRUE)

DimPlot(sc, reduction = "umap", label = TRUE, group.by = "SingleR.labels.bulk", repel = TRUE)
DimPlot(sc, reduction = "umap", label = TRUE, group.by = "SingleR.labels.sc", repel = TRUE)

DimPlot(sc, reduction = "phate", label = TRUE, group.by = "SingleR.labels.bulk", repel = TRUE)
DimPlot(sc, reduction = "phate", label = TRUE, group.by = "SingleR.labels.sc", repel = TRUE)
```

Here we calculate the proportions of each cell type compared to the sample as a whole, for both the bulk and single cell references.
```{r percents}
props <- calcProps(sc)
print(props[[1]])
print(props[[2]])
saveRDS(props, file = paste0("/Volumes/Home/Jen Jen Yeh Lab/Jack/scRNAseq/Seurat/Cell Type Proportions/Moncada_", 
                             samplename, ".Rds"))
```

# Subtyping
## Single Sample Classifier (SSC)
```{r SSC}
source("/Volumes/Home/Jen Jen Yeh Lab/Jack/scRNAseq/Old Single Cell Pipeline/Yeh_hg19mm10/SSC/create.classif.R")
load("/Volumes/Home/Jen Jen Yeh Lab/Jack/scRNAseq/Old Single Cell Pipeline/Yeh_hg19mm10/SSC/fitteds_public_2018-11-03")

classifier <- classifs[[1]]
res <- create.classif(dat = sc@assays$SCT@data,
                      classifier = classifier,
                      dec = NULL,
                      labels = NULL,
                      fit = classifier$fit)

res$subtypes <- ifelse(res$class == 1, "Basal", "Classical")
res$score <- log(res$predprob/(1 - res$predprob))

umap_df <- data.frame(Reductions(sc, slot = "umap")@cell.embeddings)
umap_df$subtypes <- res$subtypes
umap_df$score <- res$score

p1 <- ggplot(umap_df, mapping = aes(x = UMAP_1, y = UMAP_2, col = subtypes)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal()
p1

p2 <- ggplot(umap_df, mapping = aes(x = UMAP_1, y = UMAP_2, col = score)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal()
p2
```

## DECODER
The `decoderr` package provides us with the single sample deconvolution functions.
```{r DECODER}
ref_set <- "TCGA_RNAseq_PAAD"
sample_weights_unscaled <- Decon_single_sample(ref_set,
                                     sc@assays$SCT@data,
                                     "geneSymbol")
sample_weights <- Norm_PDAC_weights(sample_weights_unscaled)
sample_weights$samplename <- samplename

umap_df$bc_ratio <- sample_weights$bcRatio
umap_df$basal_tumor <- sample_weights_unscaled[, 9]
umap_df$classical_tumor <- sample_weights_unscaled[, 5]
umap_df$ActivatedStroma = sample_weights$ActivatedStroma
umap_df$NormalStroma = sample_weights$NormalStroma
umap_df$Immune = sample_weights$Immune
umap_df$Endocrine = sample_weights$Endocrine
umap_df$Exocrine = sample_weights$Exocrine

p3 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = bc_ratio)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p3

p4 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = basal_tumor)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p4

p5 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = classical_tumor)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p5

p6 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = ActivatedStroma)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p6

p7 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = NormalStroma)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p7

p8 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = Immune)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p8

p9 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = Endocrine)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p9

p10 <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = Exocrine)) +
  geom_point(size = 1, alpha = 0.6) + 
  theme_minimal() + 
  scale_colour_gradient(low = "lightblue", high = "red") +
  guides(color = guide_legend(override.aes = list(size = 3)))
p10
```

# Saving
Finally, we save the `Seurat` object with all the relevant analyses.
```{r finalsave}
saveRDS(sc, file = paste0("/Volumes/Home/Jen Jen Yeh Lab/Jack/scRNAseq/Seurat/RData Files/Moncada_",
                          samplename, "_hg38.Rds"))
```

